---
title: "SignatureFinder"
author: "Monika Lapinska"
date: "11 October 2016"
output: pdf_document
---
#Intall

```{r,message=FALSE}
library(devtools)
install_github("geneticsMiNIng/SignatureFinder")
```

It is necessary to install two additional package from bioconductor.

```{r,message=FALSE}
# source("https://bioconductor.org/biocLite.R")
# biocLite("topGO")
# biocLite("org.Hs.eg.db")
library(topGO)
library(org.Hs.eg.db)
```


#Overview
This document present an overview of the SignatureFinder package. Package 
was built to find the gene signature. It has only one function 
signature.algorithm(). One of the output result is vector with gene  
signature, the most important genes to recognize type of the cancer. 
Second possible output result is two element list which first is 
a classifier and  the second is vector with gene signature. To find gene 
signature, algorithm uses a Kruskal-Wallis test and base of characteristic 
function from Gene Ontology.


#Example
We need to download the data from PANCAN12 package. 

```{r, message=FALSE}
library(RTCGA.PANCAN12)
```
Data expression.cb1 and expression.cb2 contain expressions of 16115 genes 
for each patient. Data  clinical.cb contains patient id and description of 
his disease. 
```{r}
data("expression.cb1")
data("expression.cb2")
expression <- rbind(expression.cb1,expression.cb2)
data("clinical.cb")
```

In this example we use only three type of cancer.
```{r}
Names_of_cancers <- c("TCGA Lung Adenocarcinoma"
                      ,"TCGA Rectal Cancer"
                      ,"TCGA Bladder Cancer")
clinical <- clinical.cb[which(clinical.cb$X_cohort%in%Names_of_cancers),
                        c("sampleID","X_cohort")]
clinical$sampleID <- gsub("-",".",clinical$sampleID)
expression <- expression[,c(1,which(colnames(expression)%in%
                                       clinical$sampleID))]
```


We need to do some transformation of the data, because function 
signature.algorithm() need data.frame where columns contain genes 
and rows contain samples. 


```{r}
# delete the genes with NA
expression <- expression[!is.na(rowSums(expression[,-1])),]
# transpose original data 
n <- expression$Sample
expression_t <- as.data.frame(t(expression[,-1]))
colnames(expression_t) <- n
disease <- character()
disease[which(rownames(expression_t)%in%clinical$sampleID[which(clinical$
         X_cohort=="TCGA Lung Adenocarcinoma")])] <- "Lung"
disease[which(rownames(expression_t)%in%clinical$sampleID[which(clinical$
         X_cohort=="TCGA Rectal Cancer")])] <- "Rectal"
disease[which(rownames(expression_t)%in%clinical$sampleID[which(clinical$
         X_cohort=="TCGA Bladder Cancer")])] <- "Bladder"
expression_t <- cbind(expression_t,disease)
```

An example of the data after transformation:
```{r,echo=FALSE, message=FALSE}
# delete the genes with NA
expression_t[1:5,c(sample(1:dim(expression_t)[2]-1,5),dim(expression_t)[2])]

```

We want to show only example of usage, so we select a subset of genes. 
```{r, message=FALSE}
expression_data <- expression_t[,c(sample(1:dim(expression_t)[2],300),
                                   dim(expression_t)[2])]
```


Now we can use Signature.algorithm() to find gene signature. There is five 
diferent method to calculate gene signature. We will briefly describe 
each of them.

all.all.med - For k best genes selected using Kruskal-Wallis test choose all
characteristic function. Then for each gene we use all its fisher values 
compatible with characteristic function. Next step is to calculate median 
for k best genes and as signature accepts k best genes and other which are 
smaller then median.

```{r, message=FALSE}
SignatureFinder::signature.algorithm(expression_data,"disease",k=20,
                                     signature.method="all.all.med")
```

all.min.max - For k best genes selected using Kruskal-Wallis test choose all
characteristic function. Then for each gene we use minimal fisher values 
compatible with characteristic function. Next step is to calculate maximum 
for k best genes and as signature accepts k best genes which are smaller 
then maximum.

```{r, message=FALSE}
SignatureFinder::signature.algorithm(expression_data,"disease",k=20,
                                     signature.method="all.min.max")
```

all.5.med - For k best genes selected using Kruskal-Wallis test choose all 
characteristic function. Then for each gene we use 5 smalest its fisher 
values compatible with characteristic function. Next step is to calculate 
median for k best genes and as signature accepts k best genes and other 
which are smaller then median.

```{r, message=FALSE}
SignatureFinder::signature.algorithm(expression_data,"disease",k=20,
                                     signature.method="all.5.med")
```


m.all.max - For k best genes selected using Kruskal-Wallis test choose only 
m  characteristic function. Then for each gene we use all its fisher values 
compatible with characteristic function. Next step is to calculate maximum 
for k best genes and as signature accepts k best genes which are smaller 
then maximum.

```{r, message=FALSE}
SignatureFinder::signature.algorithm(expression_data,"disease",k=20,
                                     signature.method="m.all.max",m=15)
```

p.all.med - For k best genes selected using Kruskal-Wallis test choose only 
this  characteristic function which p-value are smaller then p. Then for 
each gene we use all its fisher values compatible with characteristic 
function. Next step is to calculate median for k best genes and as signature
accepts k best genes and other which are smaller then median.

```{r, message=FALSE}
SignatureFinder::signature.algorithm(expression_data,"disease",k=20,
                                     signature.method="p.all.med",p=0.02)
```

Second possible of output is create classifier. We can choose from 
Random Forest and SVM method. But we also have a vector of gene signature. 
In this case, the output is a list where first component is classifier 
and second component is vector with gene signature.

Again, we can choose method to calculate the signature, but this time we 
leave the default settings. Only we need to change is out parameter. 

We want to show how good is the classifier so we need to choose train and 
test data. And use function only on the training data. We show how to built 
classifier in bothmethod SVM and random forest. We use default settings in 
method to calculate the signature, hence that in both signature there are 
the same genes.

```{r, message=FALSE}
set.seed(123)
indeks_train <-sample(1:dim(expression_data)[1],dim(expression_data)[1]*2/3)
data_train <- expression_data[indeks_train,]
data_test <- expression_data[-indeks_train,]
SVM.out <- SignatureFinder::signature.algorithm(data_train,"disease",k=20,
                                             out="predictive.SVM")
RF.out <- SignatureFinder::signature.algorithm(data_train,"disease",k=20,
                                             out="predictive.RandomForest")
clasifier_SVM <- SVM.out[[1]]
clasifier_RF <- RF.out[[1]]
signature <- SVM.out[[2]]


```

Now we transform test data and then show the contingency table of the result. 


```{r, message=FALSE}
names_of_gene <- colnames(data_test)[-dim(data_test)[2]]
index_of_gene <- which(names_of_gene %in% signature)
data_test_sig <- data_test[,index_of_gene]
y <- data_test[,dim(data_test)[2]]
y <- as.factor(y)
data_test_sig <- data.frame(data_test_sig)
score_SVM <- predict(clasifier_SVM,data_test_sig)
score_RF <- predict(clasifier_RF,data_test_sig)
```

```{r,echo=FALSE, message=FALSE}
score_SVM <- as.character(score_SVM)
score_SVM[which(score_SVM=="1")] <- "Bladder"
score_SVM[which(score_SVM=="2")] <- "Lung"
score_SVM[which(score_SVM=="3")] <- "Rectal"
score_SVM <- as.factor(score_SVM)
```

Result of SVM classifier:
```{r, message=FALSE}
table(pred=score_SVM,real=y)
```

```{r,echo=FALSE, message=FALSE}
score_RF <- as.character(score_RF)
score_RF[which(score_RF=="1")] <- "Bladder"
score_RF[which(score_RF=="2")] <- "Lung"
score_RF[which(score_RF=="3")] <- "Rectal"
score_RF <- as.factor(score_RF)
```

Result of random forest classifier:
```{r, message=FALSE}
table(pred=score_RF,real=y)
```

```{r, echo=FALSE}

```
